A seamless, client-centric programming model for type safe web applications

### 2014-03-01 paper.tex
Remove besserwisser words (of course, obviously, straightforward, etc.)

The \lstinline!export! function takes an arbitrary function, provided that all
its arguments as well as its return value are serializable through the aptly
named \lstinline!Serialize! type class, and produces a typed identifier which
may be used to refer to the exported function. In this example, the type of
\lstinline!greetings! is \lstinline!Export (String -> Server ())!,
indicating that the identifier refers
to an exported function with a single \lstinline!String! argument and no return
value. Exported functions all live in the \lstinline!Server! monad.
This part of the program is executed on both the server and the client, albeit
with slightly different side effects, as described in section \ref{sec:impl}.

After the \lstinline!export! call, we enter the domain of client-exclusive code
with the application of \lstinline!runClient!. This function, unsurprisingly,
execute computations in the \lstinline!Client! monad which is essentially an
IO monad with cooperative multitasking added on top, to mitigate the fact that
Javascript has no native concurrency support. \lstinline!runClient! does not
return, and is the only function with a return type of \lstinline!App Done!,
which ensures that each \lstinline!App! computation contains exactly one client
computation.

@@

This part is entirely straightforward; the URL of the server is provided for
the benefit of the client and the port to listen on is provided for the
benefit of the server. The final argument is the main function of our
application. Figure \ref{lst:hello-server} gives the code of this example in
its entirety.

@@

The client implementation of \lstinline!export! is straightforward.

@@

As we can see, the implementation of our programming model is very
straightforward and requires no bothersome compiler modifications or language
extensions and is thus easily portable to other Haskell compilers.

@@

In order to allow the same language to be used on both client and server, we
obviously need some way to compile that language into Javascript. To this end,
we make use of the Haste compiler\ \cite{haste}, started as an MSc thesis and
continued as part of this work. Haste builds on the GHC compiler to provide
the full Haskell language, including most GHC-specific extensions, in the
browser.

@@

The runtime data representation of Haste programs is kept as close to regular
Javascript programs as possible. The numeric types are represented using the
Javascript \lstinline!Number! type, which is defined as the IEEE754 double
precision floating point type. This obviously adds some overhead to operations
on integers as overflow and non-integer divisions must be handled. However, this
is common practice in hand-written Javascript as well, and is generally handled
efficiently by Javascript engines.

Values of non-primitive data types in Haskell consist of a data constructor and
zero or more arguments. In Haste, these values are represented using arrays,
with the first element representing the data constructor and the following
values representing its arguments. For instance, the value \lstinline!42 :: Int!
is represented as \lstinline![0, 42]!, the leading \lstinline!0! representing
the zeroth constructor of the \lstinline!Int! type and the \lstinline!42!,
obviously, representing the ``machine'' integer. It may seem strange that a
limited precision integer is represented using one level of indirection rather
than as a simple number, but recall that the \lstinline!Int! type is defined by
GHC as \lstinline!data Int = I# Int#! where \lstinline!Int#! is the primitive
type for machine integers.

@@

While Haste supports the Foreign Function Interface inherited from GHC, with
its usual features and limitations \ \cite{ffi}, it is often impractical to
work within the confines of an interface designed for communication on a very
low level. For this reason Haste sports its own method for interacting with
Javascript as well, which allows the programmer to pass any value back and
forth between Haskell and Javascript, as long as she can come up with a way to
translate this value between its Haskell and Javascript representations.
Of course, not performing any translation at all is also a valid
``translation'', which allows Haskell code to store any Javascript value for
later retrieval without inspecting it and vice versa. The example given in
figure \ref{lst:ffi} implements mutable variables using this custom Javascript
interface.

@@

There are several benefits to this method, the most obvious being that new
marshallable types can be added by simply instantiating a type class. Thanks
to the lazy evaluation employed by Haste, each foreign function object is only
created once and then cached; any further calls to the same (Haskell) function
will reuse the cached function object. Implementation-wise, this method is also
very non-intrusive, requiring only the use of the normal FFI to import
Javascript's \lstinline!eval! function; no modification of the compiler is
needed.

@@

Sunroof is completely type-safe, in the DSL itself as well as in the
communication with the Haskell host. However, the fact that client and server
must be written in two separate languages - any code used to generate
Javascript must of course be built solely from the primitives of the Sunroof
language, precluding use of general Haskell code - makes code reuse hard.

@@

Our programming model is implemented as a library, assuming only two Haskell
compilers, one targeting Javascript and one targeting the programmer's server
platform of choice. While we use Haste as our Javascript-targeting compiler,
modifying our implementation to use GHCJS or even the Javascript backend of UHC
would be entirely straightforward. This implementation not only allows for
greater flexibility, but also eliminates the need to tangle with complex
compiler internals.

\paragraph{Inspiration and alternatives to \lstinline!export!} One crucial
aspect of implementing cross-network function calls is, of course, the issue of
data representation: the client side of things must be able to obtain some
representation of any function it may want to call on the server.

@@

\paragraph{Client-centricity} Unlike Conductance and Duetto, our approach
takes a firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client-side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an
\lstinline!onClient! counterpart to \lstinline!onServer! would be relatively
straightforward.

@@

\paragraph{Information flow control} Web applications often make use of a wide
range of third party code for user tracking, advertising, collecition of
statistics and a wide range of other tasks. Any piece of code executing in the
context of a particular web session may not only interact with any other piece
of code executing in the same context, but may also perform basically limitless
communication with third parties and may thus, inadvertently or not, leak
information about the application state. This is obviously highly undesirable
for many applications, which is why there is ongoing work in controlling the
information flow within web applications\ \cite{jsflow}.

>>>>

The \lstinline!export! function takes an arbitrary function, provided that all
its arguments as well as its return value are serializable through the
\lstinline!Serialize! type class, and produces a typed identifier which
may be used to refer to the exported function. In this example, the type of
\lstinline!greetings! is \lstinline!Export (String -> Server ())!,
indicating that the identifier refers
to an exported function with a single \lstinline!String! argument and no return
value. Exported functions all live in the \lstinline!Server! monad.
This part of the program is executed on both the server and the client, albeit
with slightly different side effects, as described in section \ref{sec:impl}.

After the \lstinline!export! call, we enter the domain of client-exclusive code
with the application of \lstinline!runClient!. This function, execute
computations in the \lstinline!Client! monad which is essentially an
IO monad with cooperative multitasking added on top, to mitigate the fact that
Javascript has no native concurrency support. \lstinline!runClient! does not
return, and is the only function with a return type of \lstinline!App Done!,
which ensures that each \lstinline!App! computation contains exactly one client
computation.

@@

The URL of the server is provided for the benefit of the client and the port
to listen on is provided for the benefit of the server. The final argument is
the main function of our application. Figure \ref{lst:hello-server} gives the
code of this example in its entirety.

@@

The client implementation of \lstinline!export! is not very complex.

@@

As we can see, the implementation of our programming model is rather simple
and requires no bothersome compiler modifications or language extensions,
and is thus easily portable to other Haskell compilers.

@@

In order to allow the same language to be used on both client and server, we
need some way to compile that language into Javascript. To this end, we make
use of the Haste compiler\ \cite{haste}, started as an MSc thesis and continued
as part of this work. Haste builds on the GHC compiler to provide the full
Haskell language, including most GHC-specific extensions, in the browser.

@@

The runtime data representation of Haste programs is kept as close to regular
Javascript programs as possible. The numeric types are represented using the
Javascript \lstinline!Number! type, which is defined as the IEEE754 double
precision floating point type. This adds some overhead to operations on
integers as overflow and non-integer divisions must be handled. However, this
is common practice in hand-written Javascript as well, and is generally handled
efficiently by Javascript engines.

Values of non-primitive data types in Haskell consist of a data constructor and
zero or more arguments. In Haste, these values are represented using arrays,
with the first element representing the data constructor and the following
values representing its arguments. For instance, the value \lstinline!42 :: Int!
is represented as \lstinline![0, 42]!, the leading \lstinline!0! representing
the zeroth constructor of the \lstinline!Int! type and the \lstinline!42!
representing the ``machine'' integer. It may seem strange that a limited
precision integer is represented using one level of indirection rather
than as a simple number, but recall that the \lstinline!Int! type is defined by
GHC as \lstinline!data Int = I# Int#! where \lstinline!Int#! is the primitive
type for machine integers.

@@

While Haste supports the Foreign Function Interface inherited from GHC, with
its usual features and limitations \ \cite{ffi}, it is often impractical to
work within the confines of an interface designed for communication on a very
low level. For this reason Haste sports its own method for interacting with
Javascript as well, which allows the programmer to pass any value back and
forth between Haskell and Javascript, as long as she can come up with a way to
translate this value between its Haskell and Javascript representations. Not
performing any translation at all is also a valid ``translation'', which allows
Haskell code to store any Javascript value for later retrieval without
inspecting it and vice versa. The example given in figure \ref{lst:ffi}
implements mutable variables using this custom Javascript interface.

@@

There are several benefits to this method, the most prominent being that new
marshallable types can be added by simply instantiating a type class. Thanks
to the lazy evaluation employed by Haste, each foreign function object is only
created once and then cached; any further calls to the same (Haskell) function
will reuse the cached function object. Implementation-wise, this method is also
very non-intrusive, requiring only the use of the normal FFI to import
Javascript's \lstinline!eval! function; no modification of the compiler is
needed.

@@

Sunroof is completely type-safe, in the DSL itself as well as in the
communication with the Haskell host. However, the fact that client and server
must be written in two separate languages - any code used to generate
Javascript must be built solely from the primitives of the Sunroof
language in order to be compilable into Javascript, precluding use of general
Haskell code - makes code reuse hard.

@@

Our programming model is implemented as a library, assuming only two Haskell
compilers, one targeting Javascript and one targeting the programmer's server
platform of choice. While we use Haste as our Javascript-targeting compiler,
modifying our implementation to use GHCJS or even the Javascript backend of UHC
would be trivial. This implementation not only allows for greater flexibility,
but also eliminates the need to tangle with complex compiler internals.

\paragraph{Inspiration and alternatives to \lstinline!export!} One crucial
aspect of implementing cross-network function calls is the issue of data
representation: the client side of things must be able to obtain some
representation of any function it may want to call on the server.

@@

\paragraph{Client-centricity} Unlike Conductance and Duetto, our approach
takes a firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client-side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an
\lstinline!onClient! counterpart to \lstinline!onServer! is a simple task.

@@

\paragraph{Information flow control} Web applications often make use of a wide
range of third party code for user tracking, advertising, collecition of
statistics and a wide range of other tasks. Any piece of code executing in the
context of a particular web session may not only interact with any other piece
of code executing in the same context, but may also perform basically limitless
communication with third parties and may thus, inadvertently or not, leak
information about the application state. This is of course highly undesirable
for many applications, which is why there is ongoing work in controlling the
information flow within web applications\ \cite{jsflow}.
